---
title: "p8105_hw2_nel2139"
author: "Nicole Lewis"
date: "2026-10-07"
output: github_document
---

## Load packages

```{r setup}
library(tidyverse)
library(dplyr)
library(readxl)
```

## Problem 1

```{r}
nyc_transit_df = 
  read_csv(
    "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
    na = c("", "NA"), 
     col_types = cols(
      Route8  = col_character(),
      Route9  = col_character(),
      Route10 = col_character(),
      Route11 = col_character())) |> 
  janitor::clean_names() |> 
  select(line:entry, vending, ada) |> 
  mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
nyc_transit_df
dim(nyc_transit_df)
```

This dataset contains information about the NYC subway system including the subway `line`, `station_name`, `station_latitude`, `station_longitude`, the routes served at each station (stored across `route1` through `route11`), the `entrance_type`, whether `entry` is allowed, whether there is `vending`, and whether the station is `ada` compliant. To clean the data, I imported the CSV file using a relative path and treated blank cells as missing. Since `route1` through `route7` mostly listed letter trains like the A and N trains, they were treated as character variables in R. However, `route8` through `route11` only consisted of numbered trains like the 1 train so they were treated as numeric (dbl) in R. To ensure consistency, I reclassified `route8` through `route11` from numeric (double) to character variables, so that all route columns are the same type. I then cleaned the variable names, kept only the variables listed above, and converted `entry` from a character variable with YES/NO values to a logical variable with TRUE/FALSE values. The dataset has `r nrow(nyc_transit_df)` rows and `r ncol(nyc_transit_df)` columns. These data are not tidy because route is a single variable spread across 11 columns. In tidy data, each variable should have its own column, so the routes should instead be stored in one column, with a separate row for each route served. 

Number of distinct stations

```{r}
n_stations = 
  nyc_transit_df |> 
  distinct(line, station_name) |> 
  nrow()
n_stations
```

There are `r n_stations` distinct stations. 

ADA compliant stations

```{r}
n_ada_stations = 
  nyc_transit_df |> 
  filter(ada == TRUE) |> 
  distinct(line, station_name) |> 
  nrow()
n_ada_stations
```

`r n_ada_stations` stations are ADA compliant.

Proportion of station entrances/exits without vending that allow entrance

```{r}
prop_no_vending_entry = 
  nyc_transit_df |> 
  select(vending, entry) |> 
  filter(vending == "NO") |> 
  summarize(prop = sum(entry == TRUE) / n()) 
prop_no_vending_entry
```

`r round(prop_no_vending_entry, 3)` of station entrances/exits without vending allow entrance.

Reformat the data using `pivot_longer`

```{r}
nyc_transit_long = 
  nyc_transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name"
  )
```

```{r}
n_a_stations = 
  nyc_transit_long |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()

n_a_ada_stations = 
  nyc_transit_long |>
  filter(route_name == "A", ada == TRUE) |> 
  distinct(line, station_name) |>
  nrow()
```

`r n_a_stations` distinct stations serve the A train. Of these, `r n_a_ada_stations` are ADA compliant.


## Problem 2

Mr. Trash Wheel 

```{r}
mr_trash_df = read_excel(
  "./data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Mr. Trash Wheel",
  range = "A2:N742",
  na = c("", "NA")) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.numeric(year),
    trash_wheel = "mr_trash_wheel")
```

Professor Trash Wheel 

```{r}
prof_trash_df = read_excel(
  "./data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Professor Trash Wheel",
  range = "A2:M139",
  na = c("", "NA", "dive")) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(trash_wheel = "prof_trash_wheel")
```

Gwynnda

```{r}
gwynnda_df = read_excel(
  "./data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Gwynnda the Good Wheel of the W",
  range = "A2:L394",
  na = c("", "NA")) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(trash_wheel = "gwynnda")
```

Combine datasets

```{r}
trash_wheel_df = 
  bind_rows(mr_trash_df, prof_trash_df, gwynnda_df) |> 
  select(trash_wheel, everything())
```

The combined dataset contains data collected by three trash wheels in Baltimore: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. It has `r nrow(trash_wheel_df)` observations and `r ncol(trash_wheel_df)` variables, where each observation is one dumpster of trash collected by a wheel. Key variables present in all three datasets include `trash_wheel` (which wheel collected the trash), `dumpster`, `month`, `year`, `date`, `weight_tons`, `volume_cubic_yards`, `homes_powered`, and counts of specific items such as `plastic_bottles`, `polystyrene`, `cigarette_butts`, `plastic_bags` and `wrappers`. Some items were only recorded for certain wheels: `sports_balls` is only available for Mr. Trash Wheel, and `glass_bottles` is not recorded for Gwynnda. The data span from `r min(trash_wheel_df$year)` to `r max(trash_wheel_df$year)`.

For available data, the total weight of trash collected by Professor Trash Wheel was `r sum(prof_trash_df$weight_tons)` tons. The total number of cigarette butts collected by Gwynnda in June of 2022 was `r sum(filter(gwynnda_df, month == "June", year == 2022)$cigarette_butts)`

