---
title: "p8105_hw2_nel2139"
author: "Nicole Lewis"
date: "2026-10-07"
output: github_document
---

## Load packages

```{r setup}
library(tidyverse)
library(dplyr)
library(readxl)
```

## Problem 1

```{r}
nyc_transit_df = 
  read_csv(
    "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
    na = c("", "NA"), 
     col_types = cols(
      Route8  = col_character(),
      Route9  = col_character(),
      Route10 = col_character(),
      Route11 = col_character())) |> 
  janitor::clean_names() |> 
  select(line:entry, vending, ada) |> 
  mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
nyc_transit_df
dim(nyc_transit_df)
```

This dataset contains information about the NYC subway system including the subway `line`, `station_name`, `station_latitude`, `station_longitude`, the routes served at each station (stored across `route1` through `route11`), the `entrance_type`, whether `entry` is allowed, whether there is `vending`, and whether the station is `ada` compliant. To clean the data, I imported the CSV file using a relative path and treated blank cells as missing. Since `route1` through `route7` mostly listed letter trains like the A and N trains, they were treated as character variables in R. However, `route8` through `route11` only consisted of numbered trains like the 1 train so they were treated as numeric (dbl) in R. To ensure consistency, I reclassified `route8` through `route11` from numeric (double) to character variables, so that all route columns are the same type. I then cleaned the variable names, kept only the variables listed above, and converted `entry` from a character variable with YES/NO values to a logical variable with TRUE/FALSE values. The dataset has `r nrow(nyc_transit_df)` rows and `r ncol(nyc_transit_df)` columns. These data are not tidy because route is a single variable spread across 11 columns. In tidy data, each variable should have its own column, so the routes should instead be stored in one column, with a separate row for each route served. 

Number of distinct stations

```{r}
n_stations = 
  nyc_transit_df |> 
  distinct(line, station_name) |> 
  nrow()
n_stations
```

There are `r n_stations` distinct stations. 

ADA compliant stations

```{r}
n_ada_stations = 
  nyc_transit_df |> 
  filter(ada == TRUE) |> 
  distinct(line, station_name) |> 
  nrow()
n_ada_stations
```

`r n_ada_stations` stations are ADA compliant.

Proportion of station entrances/exits without vending that allow entrance

```{r}
prop_no_vending_entry = 
  nyc_transit_df |> 
  select(vending, entry) |> 
  filter(vending == "NO") |> 
  summarize(prop = sum(entry == TRUE) / n()) 
prop_no_vending_entry
```

`r round(prop_no_vending_entry, 3)` of station entrances/exits without vending allow entrance.

Reformat the data using `pivot_longer`

```{r}
nyc_transit_long = 
  nyc_transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name"
  )
```

```{r}
n_a_stations = 
  nyc_transit_long |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()

n_a_ada_stations = 
  nyc_transit_long |>
  filter(route_name == "A", ada == TRUE) |> 
  distinct(line, station_name) |>
  nrow()
```

`r n_a_stations` distinct stations serve the A train. Of these, `r n_a_ada_stations` are ADA compliant.


## Problem 2

Mr. Trash Wheel 

```{r}
mr_trash_df = read_excel(
  "./data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Mr. Trash Wheel",
  range = "A2:N742",
  na = c("", "NA")) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.numeric(year),
    trash_wheel = "mr_trash_wheel")
```

Professor Trash Wheel 

```{r}
prof_trash_df = read_excel(
  "./data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Professor Trash Wheel",
  range = "A2:M139",
  na = c("", "NA", "dive")) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(trash_wheel = "prof_trash_wheel")
```

Gwynnda

```{r}
gwynnda_df = read_excel(
  "./data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Gwynnda the Good Wheel of the W",
  range = "A2:L394",
  na = c("", "NA")) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(trash_wheel = "gwynnda")
```

Combine datasets

```{r}
trash_wheel_df = 
  bind_rows(mr_trash_df, prof_trash_df, gwynnda_df) |> 
  select(trash_wheel, everything())
```

The combined dataset contains data collected by three trash wheels in Baltimore: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. It has `r nrow(trash_wheel_df)` observations and `r ncol(trash_wheel_df)` variables, where each observation is one dumpster of trash collected by a wheel. Key variables present in all three datasets include `trash_wheel` (which wheel collected the trash), `dumpster`, `month`, `year`, `date`, `weight_tons`, `volume_cubic_yards`, `homes_powered`, and counts of specific items such as `plastic_bottles`, `polystyrene`, `cigarette_butts`, `plastic_bags` and `wrappers`. Some items were only recorded for certain wheels: `sports_balls` is only available for Mr. Trash Wheel, and `glass_bottles` is not recorded for Gwynnda. The data span from `r min(trash_wheel_df$year)` to `r max(trash_wheel_df$year)`.

For available data, the total weight of trash collected by Professor Trash Wheel was `r sum(prof_trash_df$weight_tons)` tons. The total number of cigarette butts collected by Gwynnda in June of 2022 was `r sum(filter(gwynnda_df, month == "June", year == 2022)$cigarette_butts)`. 


## Problem 3

Baseline dataset 

```{r}
mci_baseline_df = 
  read_csv("./data/MCI_baseline.csv",
           skip = 1,
           na = c(".", "NA", "")) |> 
  janitor::clean_names() |> 
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male",
        0 ~ "female"),
    sex = as.factor(sex),
    apoe4 = 
      case_match(
        apoe4, 
        1 ~ "carrier",
        0 ~ "non-carrier"),
    apoe4 = as.factor(apoe4)) |> 
  filter(current_age < age_at_onset | is.na(age_at_onset))
```

The baseline data were imported with `read_csv()`. The first row of the datafile contains descriptions of each variable, so `skip = 1` was used to ensure this section was not included in the imported data, and that the second row was used as the column names. Participants who did not develop MCI during the follow up period have no age at onset, so their age at onset was recorded as `.` in the data file. These values were recoded to NA, which ensured that `age_at_onset` was treated as a numeric variable, rather than a character variable. Variable names were then cleaned and converted to snake case. Sex and APOE4 status were originally coded as 0 and 1 and therefore treated as numeric in R. Since these are categorical variables, they were recoded with `case_match()` as male/female and carrier/non-carrier, respectively, and converted to factor variables with `as.factor()`. Finally, `filter()` was used to remove participants who did not meet the inclusion criteria of no MCI at baseline. Participants whose `age_at_onset` was at or before their baseline `current_age` already had MCI at baseline and were removed (n=4), while participants with no age at onset were kept because they met the inclusion criteria; they were MCI-free at baseline and simply did not develop MCI during follow-up. Participants who did not develop MCI have `NA` for `age_at_onset`. This dataset contains demographic information for `r nrow(mci_baseline_df)` participants measured at the study baseline and includes `r ncol(mci_baseline_df)` variables: `id`, `current_age`, `sex`, years of `education`, `apoe4` carrier status, and `age_at_onset` of MCI. While the other variables were measured at baseline, `age_at_onset` was recorded during the follow-up period. 


```{r}
n_recruited = 
  read_csv("./data/MCI_baseline.csv", skip = 1, na = c(".", "NA", "")) |> 
  nrow()

n_eligible = nrow(mci_baseline_df)

n_excluded = n_recruited - n_eligible

n_mci = 
  mci_baseline_df |> 
  filter(!is.na(age_at_onset)) |> 
  nrow()
```

A total of `r n_recruited` participants were recruited. After excluding `r n_excluded` participants who had MCI at baseline, `r n_eligible` participants met the inclusion criteria, and `r n_mci` of these developed MCI during follow-up.

The average baseline age is `r round(mean(pull(mci_baseline_df, current_age)), 2)` years. 

```{r}
prop_women_apoe4 = 
  mci_baseline_df |> 
  filter(sex == "female") |> 
  summarize(prop = mean(apoe4 == "carrier"))
```

The proportion of women in the study who are APOE4 carriers is `r prop_women_apoe4`. 

Amyloid dataset 

```{r}

mci_amyloid_df = 
  read_csv("./data/mci_amyloid.csv",
           skip = 1,
           na = c("NA", "Na", "")) |> 
  janitor::clean_names() |> 
  rename(id = study_id) |> 
  pivot_longer(
    baseline:time_8,
    names_to = "years_since_baseline",
    names_prefix = "time_",
    values_to = "amyloid_ratio") |> 
  mutate(
    years_since_baseline = replace(years_since_baseline, years_since_baseline == "baseline", "0"), 
    years_since_baseline = as.numeric(years_since_baseline))
```

The amyloid dataset was imported with `read_csv()`. Like the baseline data, the first row of the datafile contains a description of the variables, so `skip = 1` was used to ensure this row was not included in the imported data, and that the second row was used as the column names. Missing amyloid ratio values were recorded as `NA` and `Na` in the data file, so `na = c("NA", "Na", "")` was used to read all of them as missing and ensure the amyloid ratios were classified as numeric. Variable names were then cleaned and converted to snake case and `study_id` was renamed to `id` to match the baseline dataset. The data were originally in wide format, with a separate column for the amyloid ratio at each time point. So, `pivot_longer` was used to convert the data into long format; the five time point columns were stacked into two columns, `years_since_baseline`, which records the time point, and `amyloid_ratio`, which records the measured ratio value. The time point variable was named `years_since_baseline` and treated as numeric because the description row of the datafile states that each time point is measured in years elapsed since the study baseline, which represents an amount of time which has numeric meaning. This label makes it clearer what the values represent. Then, the time point values were cleaned by removing the "time_" prefix, replacing "baseline" with 0, and converting the variable to numeric so that the `years_since_baseline` has values of 0, 2, 4, 6, and 8. This dataset contains `r n_distinct(mci_amyloid_df$id)` participants and `r nrow(mci_amyloid_df)` rows (one per participant per time point), with `r ncol(mci_amyloid_df)` variables. The variables included in this dataset are: `id`, `years_since_baseline`, and `amyloid_ratio`. The `amyloid_ratio` variable records the amyloid β 42/40 ratio, a biomarker that changes over time and has been linked to the manifestation of clinical symptoms of Alzheimer’s disease. For each participant, this ratio is recorded at baseline and at 2, 4, 6, and 8 years after baseline. However, there are `r sum(is.na(mci_amyloid_df$amyloid_ratio))` missing amyloid ratio values in total, and `r n_distinct(filter(mci_amyloid_df, is.na(amyloid_ratio))$id)` of participants have at least one missing time point. 





