---
title: "p8105_hw2_nel2139"
author: "Nicole Lewis"
date: "2026-10-07"
output: github_document
---

## Load packages

```{r setup}
library(tidyverse)
library(dplyr)
```

## Problem 1

```{r}
nyc_transit_df = 
  read_csv(
    "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
    na = c("", "NA"), 
     col_types = cols(
      Route8  = col_character(),
      Route9  = col_character(),
      Route10 = col_character(),
      Route11 = col_character())) |> 
  janitor::clean_names() |> 
  select(line:entry, vending, ada) |> 
  mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
nyc_transit_df
dim(nyc_transit_df)
```

This dataset contains information about the NYC subway system including the subway `line`, `station_name`, `station_latitude`, `station_longitude`, the routes served at each station (stored across `route1` through `route11`), the `entrance_type`, whether `entry` is allowed, whether there is `vending`, and whether the station is `ada` compliant. To clean the data, I imported the CSV file using a relative path and treated blank cells as missing. Since `route1` through `route7` mostly listed letter trains like the A and N trains, they were treated as character variables in R. However, `route8` through `route11` only consisted of numbered trains like the 1 train so they were treated as numeric (dbl) in R. To ensure consistency, I reclassified `route8` through `route11` from numeric (double) to character variables, so that all route columns are the same type. I then cleaned the variable names, kept only the variables listed above, and converted `entry` from a character variable with YES/NO values to a logical variable with TRUE/FALSE values. The dataset has `r nrow(nyc_transit_df)` rows and `r ncol(nyc_transit_df)` columns. These data are not tidy because route is a single variable spread across 11 columns. In tidy data, each variable should have its own column, so the routes should instead be stored in one column, with a separate row for each route served. 

Number of distinct stations

```{r}
n_stations = 
  nyc_transit_df |> 
  distinct(line, station_name) |> 
  nrow()
n_stations
```


There are `n_stations` distinct stations. 


